\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{url}

\begin{document}

\title{Multi Omics Integration}

\author{
\IEEEauthorblockN{Aman Kumar Singh}
\IEEEauthorblockA{
SEMTM0045 – Data Science Methods and Practice\\
University of Bristol}
}

\maketitle

\begin{abstract}
Provide a concise overview of the research problem, the identified gap, the proposed methodological direction, and the expected impact of the project.

\end{abstract}

\begin{IEEEkeywords}
keyword1, keyword2, ..
\end{IEEEkeywords}

\section{Introduction}
\footnote{GitHub link: \url{https://github.com/UoB-DSMP2026/research-portfolio-fp25098}}



\section{Literature Review}

Multi-omics integration has significantly advanced personalized medicine by allowing a more comprehensive understanding of complex biological systems beyond single-omics analyzes \cite{acharya_comprehensive_2024}. By integrating diverse modalities such as genomics, transcriptomics and epigenomics, these approaches offer deeper insight into disease mechanisms and patient heterogeneity \cite{hong_multi-omics_2025}. However, despite these advancements, the field continues to face fundamental computational, methodological, and evaluation challenges, most notably the curse of dimensionality \cite{guarino_intrinsic-dimension_2023}. Multi-omics datasets are inherently characterised by extremely high-dimensional feature spaces, coupled with relatively small sample sizes and low signal-to-noise ratios\cite{kioroglou_multi-omic_2025,hong_multi-omics_2025}. This high-dimensional, low sample size (HDLSS) setting necessitates rigorous preprocessing pipelines, including standardised normalisation, robust batch effect correction, data imputation, and effective dimensionality reduction to enable meaningful downstream analysis \cite{abdelaziz_multi-omics_2024}.\\
Within this context, two broad methodological clusters can be identified: (i) interpretable statistical and network-based approaches, and (ii) complex latent representation learning methods. Although dimensionality reduction is widely adopted to address high-dimensional multi-omics data, these methodological categories differ substantially in the extent to which they preserve biological interpretability. Several studies fall within the latter category and employ approaches that do not retain explicit feature-level representations\cite{shakyawar_icluf_nodate, hong_multi-omics_2025, yang_mdicc_2022}. For instance, Yang et al. \cite{yang_mdicc_2022} perform dimensionality reduction implicitly through MDICC via low-rank subspace learning, without constructing an explicit latent feature space, thereby limiting biological interpretability despite strong clustering performance. Similarly, Shakyawar et al. \cite{shakyawar_icluf_nodate} reduce dimensional complexity through iCluF via iterative refinement of patient–patient similarity networks, but provide limited insight into the underlying omics features contributing to the learned representations. Collectively, these methods reflect a broader trend in which improvements in clustering performance and computational efficiency are achieved at the cost of transparency and biological interpretability.\\
In contrast, classical approaches such as PCA  and NMF provide explicit feature contributions, enabling clearer biological interpretation \cite{downing_primer_nodate}. Network-based methods such as Similarity Network Fusion (SNF)\cite{cantini_benchmarking_2021} integrate multi-omics through fused similarity networks, while factor-based models such as MOFA \cite{darkaoui_integrative_2026} learn latent factors that retain partial biological interpretability. However, these methods are often limited in capturing complex nonlinear relationships inherent in biological systems. This contrast highlights a fundamental trade-off in the literature between interpretability and modelling expressiveness, with recent methods increasingly favouring predictive performance over biological transparency.\\
Beyond dimensionality reduction, batch effect correction represents an equally critical yet frequently underemphasised preprocessing step. Batch effects, arising from variations in experimental conditions, sequencing platforms, or laboratory protocols, can obscure true biological signals and introduce systematic bias into downstream analyses \cite{goh_why_2017}. Consequently, even well-designed dimensionality reduction techniques may fail when applied to uncorrected data. Despite this, several studies neglect robust batch correction strategies\cite{shakyawar_icluf_nodate, yang_mdicc_2022, hira_integrated_2021}. For example, Hira et al. \cite{hira_integrated_2021} integrate multi-source omics data (TCGA and GDC) without explicit batch correction; although normalisation and sample intersection are applied, their direct concatenation likely preserves technical variability, leading to representations influenced by batch-driven artefacts rather than biological signals.
Similarly, Yang et al. \cite{yang_mdicc_2022} focus on data harmonisation via MDICC but do not explicitly address cross-platform batch effects.
Zhang et al. \cite{zhang_integrative_2025} employ ComBat, while Kioroglou et al. \cite{kioroglou_multi-omic_2025} use covariate filtering, demonstrating improved control of non-biological variation. The omission of such steps constitutes a methodological gap, as uncorrected technical noise propagates through the analytical pipeline and compromises downstream validity.\\
At the downstream analysis stage, deep learning-based approaches constitute a second major methodological cluster in multi-omics research. Architectures such as ANNs, GNNs, GCNs, and transformer-based models are widely employed to capture complex nonlinear relationships and often outperform traditional machine learning methods in predictive tasks\cite{sartori_comprehensive_2025}. However, despite these performance gains, such models remain inherently black-box, offering limited feature-level interpretability and biological insight, thereby constraining their clinical applicability \cite{kang_roadmap_2022}. For example, Wu et al. \cite{wu_deepmoic_2024} demonstrate superior performance with DeepMoIC over classical classifiers such as SVM, Random Forest, and KNN, yet fail to provide clear attribution of the molecular features driving the model's predictions, limiting its biological relevance. Although post-hoc Explainable AI (XAI) frameworks, including SHAP, LIME, and Integrated Gradients, have been introduced to extract feature-level importance from such complex models, their adoption in multi-omics studies remains limited, and they provide only approximate explanations that do not fully resolve the inherent opacity of deep learning models \cite{noauthor_full_nodate}.\\
This limitation is further amplified by the HDLSS nature of multi-omics data. Deep learning models, which are inherently data-intensive, are prone to overfitting in such settings, leading to reduced generalisability \cite{kang_roadmap_2022}. In contrast, traditional machine learning methods—particularly ensemble approaches such as XGBoost and Random Forest—demonstrate greater robustness and provide inherent feature importance measures that enhance interpretability. Empirical comparisons reinforce this distinction: Wang et al. \cite{wang_mogonet_2021} demonstrate that XGBoost outperforms their proposed deep learning model, MOGONET, on certain datasets and achieves comparable performance on others, while Hong et al. \cite{hong_multi-omics_2025} show that LightGBMXT achieves superior predictive performance (AUC = 0.9727) compared to multiple deep learning architectures. These findings challenge the implicit assumption that deep learning is universally superior, instead suggesting that simpler models may be more appropriate in HDLSS settings.\\
Another persistent and largely unresolved challenge is the lack of generalisability and robustness of proposed models. Although several studies report high predictive performance, these results are often derived from limited validation strategies, including single-cohort analyses and insufficient cross-validation \cite{shakyawar_icluf_nodate, kioroglou_multi-omic_2025, gillenwater_multi-omics_2021}. For example, Kioroglou et al.  demonstrate the potential of multi-omics profiling but restrict their analysis to a single cohort without external validation, limiting its generalisability \cite{kioroglou_multi-omic_2025}. Similarly, Hong et al. despite reporting high performance, rely on a small dataset with a simple hold-out validation strategy, raising concerns regarding overfitting and optimistic bias \cite{hong_multi-omics_2025}. Furthermore, the lack of functional validation of identified biomarkers and limited integration across diverse omics layers undermine reproducibility and biological credibility.\\
More rigorous validation strategies provide partial solutions. Benkirane et al., for instance, evaluate their framework across multiple independent datasets, demonstrating improved robustness \cite{benkirane_customics_2023}. Similarly, Wu et al. validate their framework, DeepMoIC, across multiple TCGA cohorts (e.g., BRCA, LGG, KIPAN), offering broader evaluation \cite{wu_deepmoic_2024}. However, even such approaches remain limited, as validation within a single data repository (e.g., TCGA) may not capture real-world heterogeneity due to shared batch effects, population biases, and experimental protocols. Thus, while multi-cohort validation represents progress, it does not fully resolve the challenge of generalisability.\\
Collectively, these findings reveal a critical disconnect in current multi-omics research. While advances in integration and predictive modelling have improved performance, they are not matched by equivalent progress in interpretability and rigorous validation. This imbalance limits clinical translatability, as effective personalised medicine requires not only accurate predictions but also biologically interpretable and reproducible insights. Addressing these interconnected limitations—particularly the trade-offs between performance, interpretability, and generalisability—remains a central challenge and defines a clear research gap for future work.
\section {Research Gap and Objectives}
Despite substantial progress in multi-omics integration, current approaches exhibit a fundamental imbalance between predictive performance, interpretability, and generalisability. As identified in the preceding literature, many frameworks achieve high accuracy but rely on complex latent representation learning methods that fail to retain explicit feature-level information, thereby limiting biological interpretability. This issue is further compounded by inadequate preprocessing strategies, particularly the omission of robust batch effect correction and systematic handling of missing data, which introduces technical bias and compromises downstream validity.\\
Moreover, the high-dimensional, low sample size (HDLSS) nature of multi-omics data poses significant challenges for model robustness. Deep learning-based approaches, while powerful in capturing nonlinear relationships, are prone to overfitting and often lack transparency, restricting their clinical applicability. In parallel, although traditional machine learning methods offer improved interpretability and robustness, they are not consistently leveraged within integrative frameworks.\\
A further limitation lies in the evaluation of existing models, where performance is frequently assessed using limited validation strategies, such as single-cohort analyses or simple hold-out splits. The lack of rigorous external validation across heterogeneous datasets raises concerns regarding generalisability and reproducibility. Consequently, there exists a clear need for a unified multi-omics integration framework that systematically addresses preprocessing deficiencies, balances predictive performance with interpretability, and ensures robust validation for clinically meaningful biomarker discovery.\\
\textbf{Research Question:} 
How can multi-omics integration frameworks be designed to ensure robust preprocessing, preserve feature-level interpretability, and achieve generalisable predictive performance for clinically meaningful biomarker discovery?

\textbf{Objectives:}
\begin{itemize}
    \item Develop a robust preprocessing pipeline incorporating standardisation, missing data imputation, and batch effect correction to minimise technical variability across multi-omics datasets.
    \item Implement interpretable dimensionality reduction and integration strategies that retain feature-level contributions while effectively reducing data complexity.
    \item Design a predictive modelling framework integrating multi-omics data, complemented by explainability techniques (e.g., SHAP) to identify biologically meaningful features driving predictions.
    \item Evaluate the proposed framework using rigorous validation strategies, including cross-validation and testing on independent or heterogeneous datasets, to ensure robustness and generalisability.
\end{itemize}

\section{Proposed Methodology}
To systematically address the imbalance between predictive performance, feature-level interpretability, and cross-platform generalisability identified in current multi-omics research, this study proposes a structured and integrative analytical framework. The framework begins with rigorous data preprocessing, incorporating standardisation, missing data imputation, and batch effect correction to minimise technical variability and ensure data consistency. It then applies interpretable dimensionality reduction and integration strategies to reduce data complexity while preserving biologically meaningful feature contributions. Building on these representations, robust predictive modelling approaches are employed alongside explainability techniques to enable transparent identification of key molecular drivers. Finally, the framework incorporates stringent validation procedures, including cross-validation and evaluation on independent or heterogeneous cohorts, to ensure that the resulting models are both reliable and generalisable across diverse datasets.\\
The technical pipeline of the proposed framework is presented below in a stepwise manner.
\subsubsection{Data Collection and Preprocessing} High-dimensional multi-omics datasets are obtained from globally curated repositories, including TCGA, ICGC, and NCBI GEO, enabling cross-cohort analysis across diverse cancer types. To ensure consistency while preserving cohort-specific characteristics, each dataset is processed independently through parallel preprocessing pipelines. Within each cohort, modality-specific normalisation techniques are applied (e.g., TPM/CPM scaling for RNA-Seq and Noob normalisation for DNA methylation) to harmonise measurement scales. Missing values are subsequently handled using robust non-parametric imputation methods, such as MissForest or KNN imputation, ensuring minimal information loss and maintaining data integrity for downstream analysis.
\subsubsection{Leave-One-Cohort-Out (LOCO) Strategy} To prevent data leakage and ensure a fair evaluation, a Leave-One-Cohort-Out (LOCO) strategy is applied at an early stage of the pipeline. In this approach, one entire dataset (for example, GEO) is set aside as an independent test cohort and is not used during model development. All preprocessing steps—such as batch effect correction, feature selection, and data integration—are performed only on the remaining training cohorts (e.g., TCGA and ICGC). The trained model is then evaluated on the held-out cohort, providing a realistic assessment of how well it generalises to completely unseen data from a different source.
\subsubsection{Horizontal Integration and Batch Correction} To increase sample size and partially mitigate the \( p \gg n \) problem, patient data from multiple cohorts are horizontally integrated by stacking samples across datasets. Following integration, modality-specific batch effect correction is applied to address cross-platform technical variability. \textit{ComBat-seq}, standard \textit{ComBat}, and \textit{limma}-based linear models are used depending on the data modality (sequencing, methylation, and proteomics, respectively). This ensures that batch effects are effectively reduced while preserving biologically meaningful variation across modalities.
\subsubsection{Feature Selection and Multi-Omics Integration} To reduce computational complexity and avoid supervised bias, each modality is independently filtered using unsupervised variance-based criteria, with features ranked via Median Absolute Deviation (MAD) and Coefficient of Variation (CV), retaining only the most variable features for downstream analysis. The resulting filtered datasets are then integrated and compressed using MOFA+ or Joint Non-Negative Matrix Factorisation (Joint NMF), yielding a low-dimensional latent representation that captures shared structure across omics layers.
\subsubsection{Multi-Architecture Benchmarking} The fused latent representation is evaluated using a unified benchmarking framework. L1-regularised logistic regression is used as the baseline model, and compared against classical methods (KNN, linear and RBF SVM), ensemble models (Random Forest, XGBoost, LightGBM), and deep learning models (ANN, GCN, and transformer-based architectures). This enables systematic comparison across increasing model complexity.
\subsubsection{Nested Clinical Validation Framework} The models are evaluated using a nested validation strategy. Internal validation is performed using a Leave-One-Cohort-Out (LOCO) approach to generate robust performance estimates and guide model selection. External validation is subsequently conducted on completely independent cohorts, including datasets from cBioPortal (MSK-IMPACT Clinical Sequencing) and the TARGET project, using inference-only testing to assess true real-world generalisability.
\subsubsection{Translational Interpretability and Visual Validation} To ensure biological interpretability of the learned representations, low-dimensional embeddings are visualised using UMAP and t-SNE to assess clustering and tissue-specific separation. Feature-level explanations are obtained using SHAP-based methods (TreeSHAP for ensemble models and KernelSHAP/DeepSHAP for neural networks). These attributions are mapped to MOFA+ loading matrices to trace key latent factors back to original molecular features for biomarker identification. Finally, pathway analysis using GSEA and KEGG mapping is used to validate biological relevance and alignment with known disease mechanisms.
\\


\section{Data, Ethics, and Practical Considerations}
\section{Project plan}
\section{Non-Technical Summary}

\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
